### 2024-09-19 docs/manuscript.tex
Edited manuscript text for clarity and brevity.

\PaperTitle{A Bioinformatician, Computer Scientist, and Geneticist lead bioinformatic tool development - which one is better?} % Article
                                                                                                                               % title

@@

\Keywords{} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

%----------------------------------------------------------------------------------------
%       ABSTRACT
%----------------------------------------------------------------------------------------

% The abstract could benefit from a clearer statement of the research question, methodology, and key findings. It currently mixes multiple ideas without a clear flow.
\Abstract{ The development of accurate bioinformatic software tools is
  crucial for the effective analysis of complex biological data. This
  study examines the relationship between the academic department
  affiliations of authors and the accuracy of the bioinformatic tools
  they develop. By analyzing a corpus of previously benchmarked
  bioinformatic software tools, we mapped bioinformatic tools to the
  academic fields of the corresponding authors and evaluated tool
  accuracy by field. Our results suggest that ``Medical Informatics''
  outperforms all other fields in bioinformatic software accuracy,
  with a mean proportion of wins in accuracy rankings exceeding the
  null expectation. In contrast, tools developed by authors affiliated
  with "Bioinformatics" and "Engineering" fields tend to be less
  accurate. However, after correcting for multiple testing, none of
  the results are statistically significant ($p>0.05$). Our findings
  reveal no strong association between academic field and
  bioinformatic software accuracy. These findings suggest that the
  development of interdisciplinary software applications can be
  effectively undertaken by any department with sufficient resources
  and training.  }

@@

% The introduction lacks a strong hook or clear statement of the problem being addressed. It would be helpful to more explicitly state the importance of understanding the relationship between academic department affiliation and bioinformatics tool accuracy.

Much is made of departmental divisions within academia; These can
denote research and teaching expertise \cite{ben2016s}, influence
hiring decisions, access to funding, publishing and the training of
students recruited for research projects
\cite{bourke1998institutions}. However, interdisciplinary subjects
such as bioinformatics break down the traditional barriers between
departments and subject areas
\cite{Ouzounis:2003,Eddy:2005,hogeweg2011roots}.

Bioinformatics, is an interdisciplinary field that fuses biology,
computer science, and mathematics, and now plays a pivotal role in
modern biological research \cite{Ouzounis:2003}. The development of
bioinformatic tools and software is critical for interpreting complex
biological questions, such as what evolutionary, structural and
functional analyses of genomic, transcriptomic, and proteomic data can
tell us. The field of ``bioinformatics'' can include many overlapping
research fields that include computational biology, biomathematics,
biostatistics, medical informatics and other similar areas.

Bioinformatics gained significant traction with the advent of
high-throughput sequencing technologies, which created a need for
robust computational tools to manage the large volumes of data
generated \cite{Ouzounis:2003,hogeweg2011roots}. In response,
departments specializing in bioinformatics began to emerge from
biology, computer science, and engineering faculties, each bringing
unique contributions to the field's development and expansion.

Bioinformatics integrates methods from various disciplines, fostering
novel insights that are unattainable within a single field
\cite{mazzocchi2019scientific}. This requires a deep understanding of
biological sciences for interpreting data, and strong computational
skills are required to develop algorithms and analyze the results at
scale.

Experts in \textbf{biological and health sciences}, such as genetics
and molecular biology, can ensure that software tools are biologically
relevant and accurate by identifying critical questions and guiding
computational tool development. However, biology departments may lack
the advanced computational expertise required for developing
sophisticated software. In our analysis, we have grouped the
biological and health fields under the category of ``\textbf{domain
  experts}''.

The \textbf{mathematical, engineering and computational sciences},
referred to here as the ``\textbf{development experts}'', contribute
significantly by bringing expertise in algorithm development,
mathematical modelling, statistics, and software engineering
principles. These skills are essential for creating efficient,
scalable, and robust bioinformatic tools.  However, the challenge for
development experts is gaining a deep understanding of biological domains,
which is necessary to ensure software tools are both relevant
and accurate.

Departmental differences can influence the development of
bioinformatic software tools because they reflect the varying
expertise, resources, and perspectives that different academic fields
offer. Development experts may excel in areas such as algorithm
efficiency, mathematical modeling, data handling, and software
engineering, while domain experts may bring a deep understanding of
biological questions, data interpretation, limitations, and
potentially better curation of control datasets. Consequently, the
success of a tool may rely more on the integration of diverse
forms of expertise rather than the specific departmental affiliation
of its developers. This blending of skills can mitigate potential
disparities between departments, resulting in comparable outcomes
regardless of the department of origin.

The primary objective of this study is to investigate whether the
academic department affiliation of a corresponding author is
associated with the accuracy of the bioinformatic software tools they
develop. We aim to determine if tools developed by authors from
certain academic fields, such as ``Bioinformatics'', ``Computer
Science'' or ``Genetics'' show higher or lower accuracy compared to
those from other fields. To achieve this, we analyzed a benchmarked
corpus of bioinformatic software tools and evaluated their accuracy
based on the academic affiliations of their developers.

\begin{figure*}[ht!]
\begin{center}
  \includegraphics[width=0.6\textwidth]{upset-plots.pdf}
\end{center}
  \caption{ The number and intersection of general (\textbf{A}) or
  specific (\textbf{B}) fields that have contributed to bioinformatic
  software tools included in recent benchmark studies.  The number and
  intersection of specific fields.  }
\label{fig:fig1}
\end{figure*}

\section*{Results}

We are interested in exploring the relationship between the accuracy
of bioinformatic software tools and the academic fields of study of
their developers. Using a published corpus of accuracy rankings
for bioinformatic software tools, we mapped these tools to their
respective academic fields and evaluated how software accuracy
correlates with the developers' academic affiliations.

We used a published supplement \cite{gardner2024} to obtain previously
benchmarked accuracy rankings and mapped corresponding authors'
addresses to a standardized list of specific 'fields of study'
\cite{fields2014}. These were further grouped into broader general
fields and areas of expertise. Figure~\ref{fig:fig1} illustrates the
number of tools representing each general and specific field ($N\ge 10$). Most
bioinformatic software is developed by corresponding authors who list
Genetics, Bioinformatics, Computer Science, or similar departments as
their primary address (Figure~\ref{fig:fig1}A). Among the general
fields, the Biological Sciences produce the most software tools,
followed by the Computer Sciences (Figure~\ref{fig:fig1}B).

\begin{figure*}[ht!]
\begin{center}
  \includegraphics[width=0.8\textwidth]{forest-z-Plot.pdf}
\end{center}
\caption{(\textbf{A}) A forest plot, illustrating the mean and 95\%
  confidence intervals of the proportion of times software tools
  published by a given field ``win'' in pairwise
  comparisons. Confidence intervals and the mean was determined using
  a bootstrapping procedure. Within each field the entries have been
  sorted by the mean number of wins. The sample size for each field is
  indicated by the column of numbers on the right of the figure.
  (\textbf{B}) A Z-score was computed for each distribution of
  bootstrap samples for each field. The expected proportion of wins
  for randomly selected groups of tools was used as ``$x$''
  (i.e. null=0.5).}
\label{fig:fig2}
\end{figure*}

The mean proportion of wins (e.g., if tool `A' outranks tool `B', this
counts as a win for tool `A') and the corresponding Z-scores provide a
method to rank fields of study based on the relative accuracy of
bioinformatic software benchmarked between 2005 and 2020, compared to
the expected number of wins for random groupings (i.e., $wins =
0.5$). A higher proportion of wins and a greater $(-1) * Z-score$
indicate that tools from a particular department more frequently
outperformed competing tools in independent benchmarks.

The specific and general field that outperformed all others is
``Medical Informatics'', a branch of ``Technologies'', with 16
software tools categorized under both fields. Among these, five are
different parameter options for the MAFFT multiple sequence alignment
tool \cite{katoh2008recent}.  These Medical Informatic tools have a mean proportion of wins of
0.70 and 95\% confidence interval of $0.53-0.85$, which excludes the
null value of 0.5. The corresponding Z-score is $-1.88$, and after correcting for multiple testing, the P-value is $0.29$.

At the other end of the spectrum is ``Bioinformatics''. Ironically,
authors that list ``Bioinformatics'' in their address field tend to
produce less accurate software for bioinformatic applications. The
mean proportion of wins was 0.43, with a 95\% confidence interval of
$0.33-0.53$. The corresponding Z-score is $1.20$ and after correcting
for multiple testing the P-value is $0.46$.

The general field of ``Engineering'' also ranked low, with a mean
proportion of wins of just 0.34 and a 95\% confidence interval of
$0.16-0.60$. The corresponding Z-score is $1.25$. This general field
comprises several smaller specific fields, such as "Bioengineering and
Biomedical Engineering," "Computer Engineering," and "Electrical and
Electronics Engineering." However, since none of these specific fields
individually had more than ten corresponding software tools, they were
excluded from the more detailed analysis.

The remaining general and specific academic fields have confidence
intervals that include the null value of 0.5, and relatively modest
Z-scores that range from -0.49 to 0.96. The P-values for each were
greater than $0.05$.

For the highest level field classifications of software development
expert, biological domain expert or interdisciplinary expert each had
similar mean proportions of wins (0.51, 0.49 and 0.46
respectively). The interdisciplinary experts had a lower Z-score of
$-0.87$, which corresponds to $P=0.46$ after correcting for multiple
testing.

@@

We tested the assumption that the speciality of academic departments
reflect the quality of the research software they produce.  After
correcting for multiple testing, we found no significant association
between academic expertise type and the accuracy of bioinformatic
tools, suggesting that academic department affiliation does not
correlate with software quality. Similarly, both general and specific
research fields showed no significant associations
(Figure~\ref{fig:fig2}).

Our earlier paper found that a long-term commitment to keeping
software updated was the primary factor associated with accurate
software tools. This current study complements that finding by demonstrating 
that citation metrics (e.g. journal impact factors and
author H-index), tool age, tool speed, and now academic fields of
inquiry are \textbf{not} associated with software accuracy.

We focus on software tool accuracy here \cite{weber2019essential}. While speed, usability and
some features of software tools are important, in our opinion the
primary concern for bioinformatic software tools is the accuracy of
the results they produce. As poor predictions may have long-term consequences for
our general research field.

Medical Informatics, under the broader category of "Technologies," is
identified as the top-performing group in developing accurate
bioinformatic software tools. The tools include a number of methods
for structural variation detection, single-cell profiling, long-read
assembly, multiple sequence alignment and are derived from several
different research teams.

Bioinformatics and Engineering ranked lower in terms of software
accuracy. Tools developed by authors who affiliated with
"Bioinformatics" typically had slightly lower accuracy than that of
other fields. However, this was not a statistically significant
finding. In addition (further bad news for this author),
``Biochemistry'' was similarly ranked, again this was not a
statistically significant finding.

This leads us to conclude that an individual's host department are not
reflective of the quality of software that they are capable of
producing ($p>0.05$ in all instances). As a consequence, the academic
department should not be used as a proxy for judging the potential of
software development projects.

\textbf{Limitations:} Some benchmarks rank multiple tool options,
which can introduce modest to large effects as these options may not
be independent. Additionally, the accuracy metrics used are diverse,
and some may be flawed under certain conditions; for example,
``accuracy'' can be misleading with large class-imbalances
\cite{luque2019impact}, and the N50 metric for sequence assembly has
been criticized by some commentators \cite{xie2021pdr}. Moreover, some
benchmarks are relatively small, so even small changes in rank can
have a substantial impact on the proportion of wins for intermediate
rankings. Lastly, the cohort of benchmarks has not been updated to
include more recent results.

There is often a disconnect between a developer’s training and their
departmental affiliation, as evidenced by this author whose diverse
academic and professional background spans mathematics,
bioinformatics, computer science, molecular biology, and genomics, yet
currently lists their affiliation with a Biochemistry Department. The
``Biochemistry'' departmental label does not accurately reflect the
author's training or recent publications. In fact, the author admits
to not recalling key details of glycolysis, the citric acid cycle, or
the structures of amino acids and proteins.

The last, or corresponding, author is typically the principal
investigator who leads a project. They may have limited direct
involvement in the development of any software tool, while
their role is mainly to provide resources and overall direction for
the project. However, there is likely to be a significant overlap between the
department of the first author, who is usually the primary developer
of the tool, and the department of the last author (though this was
not tested in this study). Therefore, we expect that the results will
be broadly similar if first-author departments were analyzed instead.

In conclusion, this study does not find strong evidence that the
academic department affiliation of authors is associated with the
accuracy of bioinformatic software tools. This challenges the
assumption that departmental expertise correlates with
software quality. Future research could explore other factors, such
as the nature of interdisciplinary collaborations or the training
of developers, to better understand what factors contribute to
the development of high-quality bioinformatic tools.

%Open questions: what impact does training have on improving software quality?
%We could end with echoing the call for more training and a continued 
%e.g. \cite{johanson2018software,siepel2019challenges}

\section*{Methods}

~~~~\textbf{Pre-registration:} This study's desired sample size, included variables, hypotheses, and
planned analyses were pre-registered on the Open Science Framework
%(https://doi.org/10.17605/OSF.IO/92PTZ)
prior to any unpublished data being collected \cite{gardner2024}.

@@

%The methodology section should provide more detail about how the software tools were selected and mapped to academic fields. It’s important to describe how biases or inconsistencies in mapping were addressed.
\textbf{Mapping tools to academic field:} For each software tool, the
corresponding publication(s) were identified, and the addresses of the
primary corresponding author were manually extracted when
available. If an author listed multiple addresses, only the first two
were used. In cases with multiple corresponding authors, the last
corresponding author was chosen.

@@

\textbf{Statistical analysis:} The accuracy data is derived from
benchmarks using a diverse number of metrics that include sensitivity,
specificity, PPV, FDR, error rates, AUROC, MCC and others
\cite{weber2019essential}. The number of tools ranked in any benchmark
ranged from 3 to 50. In order to obtain a representative measure of
accuracy for a field that accounts for the diversity in accuracy measures and
number of ranked tools, we employed a rank-based and
bootstrapping strategy.
We randomly sampled, with replacement, sets of 200 tools from the total of 498 tools. For each tool, a corresponding benchmark was selected at random, and the number of times the tool ``won'' against another tool was recorded, along with the total number of pairwise comparisons made. These counts of wins and total comparisons were then assigned to the corresponding specific, general, and expertise areas. This process was repeated 1,000 times to estimate the mean proportions of wins for each field, along with a 95\% confidence interval for these values (Figure~\ref{fig:fig2}A).
Additionally, we calculated a Z-score for each field to determine the number of standard deviations the mean number of wins deviates from the expected null value of 0.5 for randomly grouped tools
(Figure~\ref{fig:fig2}B).

@@

\noindent\textbf{Data and analysis scripts availability}

\noindent\textbf{Funding:} This research was supported by the MBIE data
science platform ``Beyond Prediction: explanatory and transparent data
science'', Genomics Aotearoa, and Marsden
Grants 19-UOO-040 and 20-UOA-282.

\noindent\textbf{Authors' contributions:} this work was conceived by 
PPG, designed by PPG, the analysis was carried out by PPG, interpretation of the results was by PPG,
the manuscript was drafted by  PPG.
All authors revised and edited the final manuscript. All authors approve the submitted version.

%----------------------------------------------------------------------------------------
%       REFERENCE LIST
%----------------------------------------------------------------------------------------
\bibliographystyle{unsrt}
\bibliography{references.bib}

>>>>

\PaperTitle{A Bioinformatician, Computer Scientist, and Geneticist lead bioinformatic tool development - which one is better?}

@@

\Keywords{
Bioinformatics,
Software Tool Development, 
Interdisciplinary Research, 
Departmental Affiliation, 
Software Accuracy, 
Medical Informatics, 
Academic Expertise, 
Benchmarking,
Computational Biology
} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

%----------------------------------------------------------------------------------------
%       ABSTRACT
%----------------------------------------------------------------------------------------
\Abstract{The development of accurate bioinformatic software tools is
  crucial for the effective analysis of complex biological data. This
  study examines the relationship between the academic department
  affiliations of authors and the accuracy of the bioinformatic tools
  they develop. By analyzing a corpus of previously benchmarked
  bioinformatic software tools, we mapped bioinformatic tools to the
  academic fields of the corresponding authors and evaluated tool
  accuracy by field. Our results suggest that ``Medical Informatics''
  outperforms all other fields in bioinformatic software accuracy,
  with a mean proportion of wins in accuracy rankings exceeding the
  null expectation. In contrast, tools developed by authors affiliated
  with "Bioinformatics" and "Engineering" fields tend to be less
  accurate. However, after correcting for multiple testing, no
  result is statistically significant ($p>0.05$). Our findings
  reveal no strong association between academic field and
  bioinformatic software accuracy. These findings suggest that the
  development of interdisciplinary software applications can be
  effectively undertaken by any department with sufficient resources
  and training.}

@@

Departmental divisions in academia signify research expertise,
influence hiring decisions, and impact access to funding, publishing
opportunities, and student training
\cite{bourke1998institutions,ben2016s}. However, interdisciplinary
fields such as bioinformatics blur traditional boundaries by
integrating data and methods from biology, computer science, and
mathematics to address complex research challenges
\cite{Ouzounis:2003,Eddy:2005,hogeweg2011roots}.

Bioinformatics has become essential to modern biological research,
facilitating evolutionary, structural, and functional analyses of
genomic, transcriptomic, and proteomic data
\cite{clement2018data,gauthier2019brief}. The development of accurate and
scalable bioinformatic tools is critical for interpreting such large
datasets. Which requires both biological insight and advanced
computational skills to build algorithms. The advent of
high-throughput technologies has driven the growth of bioinformatics,
leading to the establishment of specialized groups within biology,
computer science, and engineering faculties, each contributing to the
field’s expansion
\cite{hogeweg2011roots,gauthier2019brief}.

The contributions of ``domain experts'' to bioinformatics from the
biological and health sciences, such as genetics and molecular biology
are essential as they ensure software tools are relevant and accurate.
However, domain experts may lack the advanced computational expertise
needed to develop sophisticated software. In contrast, fields like
mathematics, engineering, and computational sciences — referred to
here as ``development experts'' — offer expertise in algorithm
development, mathematical modeling, statistics, and software
engineering, essential for creating efficient and scalable
bioinformatic tools.

It is possible that departmental differences may influence
bioinformatic tool development by reflecting the distinct expertise,
resources, and perspectives offered by different academic
fields. Development experts excel in computational efficiency, while
domain experts provide essential biological insights. Therefore, the
success of a tool may depend on the integration of diverse
skills rather than on the specific departmental affiliation of its
developers.

The primary objective of this study is to examine whether the academic
department affiliation of a corresponding author has a discernible
outcome on the accuracy (i.e. correctness of predictions) of the
bioinformatic tools they develop. Specifically, we aim to determine
whether tools created by authors from domain-expert,
development-expert of interdisciplinary fields differ in accuracy. To
address this, we analyzed a benchmarked corpus of bioinformatic
software tools and evaluated their accuracy based on the developers'
academic affiliations.

\begin{figure*}[t!]
  \centering
  \includegraphics[width=0.75\textwidth]{upset-plots.pdf}
  \caption{ The number and intersection of general (\textbf{A}) or
  specific (\textbf{B}) fields that have contributed to bioinformatic
  software tools included in recent benchmark studies.  The number and
  intersection of specific fields.  }
\label{fig:fig1}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

We explored the relationship between the accuracy of bioinformatic
software tools and the academic fields of their developers. Using a
previously published corpus of benchmarked accuracy rankings
\cite{gardner2024}, we mapped corresponding authors' addresses to
standardized ``fields of study'' \cite{fields2014} and grouped them into
broader categories.

%
Figure~\ref{fig:fig1} shows the number of tools for the general and
specific fields ($N\ge 10$). Most bioinformatic tools were developed
by authors affiliated with Genetics, Bioinformatics, Computer Science,
or similar departments. Among the general fields, Biological Sciences
produced the most software tools, followed by Computer Sciences.

\begin{figure*}[ht!]
\begin{center}
  \includegraphics[width=0.65\textwidth]{forest-z-Plot.pdf}
\end{center}
\caption{(\textbf{A}) A forest plot, illustrating the mean and 95\%
  confidence intervals of the proportion of times software tools
  published by a given field ``win'' in pairwise
  comparisons. Confidence intervals and the mean was determined using
  a bootstrapping procedure. Within each field the entries have been
  sorted by the mean number of wins. The sample size for each field is
  indicated by the column of numbers on the right of the figure.
  (\textbf{B}) A Z-score was computed for each distribution of
  bootstrap samples for each field. The expected proportion of wins
  for randomly selected groups of tools was used as ``$x$''
  (i.e. null=0.5).}
\label{fig:fig2}
\end{figure*}

We ranked fields based on the mean proportion of ``wins'' (i.e., for field '1' when its tool
'A' outperforms tool 'B' in benchmark 'X') and calculated Z-scores
to compare them against random expectations (i.e., $wins = 0.5$). A
higher proportion of wins and lower $(-1) * Z-score$ indicate better
overall accuracy.

``Medical Informatics'', a branch of ``Technologies'', outperformed
other fields, with a mean win proportion of 0.70 (95\% CI:
$0.53-0.85$) and a Z-score of $-1.88$. However, P = 0.29 after
multiple testing correction). Notably, this category includes five
different parameter options for the MAFFT sequence alignment tool
\cite{katoh2008recent}, and a further four separate corresponding
authors that list either the Department or Center of Biomedical
Informatics, Harvard Medical School as their affiliation
\cite{kim2010rsw,yang2013diverse,kharchenko2014bayesian,ruan2020fast}. This
and some other redundancies leaves just eight departments
representing the medical informatics field here.

In contrast, ``Bioinformatics'' has the lowest rank, with a mean win
proportion of 0.43 (95\% CI: $0.33-0.53$) and a Z-score of $1.20$ (P =
0.46 after correction). Similarly, ``Engineering'' ranked low, with a
win proportion of 0.34 (95\% CI: $0.16-0.60$) and a Z-score of $1.25$.

Other fields showed confidence intervals that included the null value
of 0.5, with modest Z-scores ranging from -0.49 to 0.96, and P-values
greater than 0.05.

When grouped by expertise type—software development experts,
biological domain experts, or interdisciplinary experts—all categories
had similar win proportions (0.51, 0.49, and 0.46,
respectively). Interdisciplinary experts had the lowest Z-score of
$-0.87$ (P = 0.46 after correction).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We tested the assumption that academic department specialization
reflects the quality of research software. After correcting for
multiple testing, we found no significant association between academic
expertise and the accuracy of bioinformatic tools. This suggests that
department affiliation does not correlate with software quality, and
neither general nor specific research fields showed any significant
links to tool accuracy.

A previous study indicated that maintaining software was the key
factor in producing accurate tools, while citation metrics, tool age,
speed are not associated with software accuracy \cite{Gardner:2022}.
Our findings complement this by showing that academic field is also not
associated with software accuracy.

While other aspects of bioinformatic tools, such as speed and
usability, are important, we emphasize that accuracy should remain the
top priority, as poor predictions can have long-term impacts on
research \cite{weber2019essential}.

Medical Informatics was the top-performing field in developing
accurate tools, these include methods for structural variation
detection, single-cell profiling, long-read assembly, multiple
sequence alignment and are derived from a limited number of research
teams. However, tools from Bioinformatics and Engineering
ranked lower, though these differences were not statistically
significant.

Therefore, an individual’s department is not a reliable indicator of the
quality of the software they produce. Academic affiliation should not
be used as a proxy for assessing the potential success of software
development projects.

\textbf{Limitations:} Some benchmarks include multiple tool options,
potentially introducing non-independent effects. The accuracy metrics
are diverse, with some limitations (e.g., issues with "accuracy" in
class-imbalanced datasets \cite{luque2019impact} and criticisms of the
N50 metric for sequence assembly \cite{xie2021pdr}). Additionally,
smaller benchmarks and smaller fields may exaggerate rank shifts. We
mitigate this in part by only considering departments with 10 or more
corresponding tools.

%Also, our data set has not been updated with more recent
%tools.

There may be little connection between a researcher’s training and
their listed department, as illustrated by this author’s background
which began in mathematics, before taking positions in departments of
computer science, molecular biology and bioinformatics, and now is
affiliated with a Biochemistry Department.

The corresponding (last) author, is typically the principal investigator and
may not be the primary tool developer, but rather oversees the
project. There is likely to be a strong overlap between the departments of the first
and last authors, but this was not explored in the current study.

\textbf{Final words:} This study does not find strong evidence linking
academic department affiliation with bioinformatic software
accuracy ($p>0.05$ in all instances). Future research should investigate other factors, such as
interdisciplinary collaborations and developer training, to understand
what drives high-quality tool development. Addressing potential biases
against interdisciplinary work \cite{bromham2016interdisciplinary} and ensuring long-term support for
essential software infrastructure will also be critical for advancing the
field \cite{siepel2019challenges}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Methods}

~~~~\textbf{Pre-registration:} The methods for this study followed the pre-registered proposal outlined
prior to any unpublished data collection \cite{gardner2024}.
%(https://doi.org/10.17605/OSF.IO/92PTZ)

@@

%The methodology section should provide more detail about how the
%software tools were selected and mapped to academic fields. It’s
%important to describe how biases or inconsistencies in mapping were
%addressed.

\textbf{Mapping tools to academic field:} For each software tool, the
corresponding publication(s) were identified, and the addresses of the
primary corresponding author were manually extracted when
available. If an author listed multiple addresses, only the first two
were used. In cases with multiple corresponding authors, the last
corresponding author was chosen.

@@

\textbf{Statistical analysis:} The accuracy data is derived from
benchmarks using a diverse number of metrics that include sensitivity,
specificity, PPV, FDR, error rates, AUROC, MCC and others
\cite{weber2019essential}. The number of tools ranked in any benchmark
ranged from 3 to 50. In order to obtain a representative measure of
accuracy for a field that accounts for the diversity in accuracy
measures and number of ranked tools, we employed a rank-based and
bootstrapping strategy.  We randomly sampled, with replacement, sets
of 200 tools from the total of 498 tools. For each tool, a
corresponding benchmark was selected at random, and the number of
times the tool ``won'' against another tool was recorded, along with
the total number of pairwise comparisons made. These counts of wins
and total comparisons were then assigned to the corresponding
specific and general departments, and expertise areas.
In other words, a tool ranked second in a benchmark of 11 tools will
contribute 9 wins and 10 comparisons to the totals for its
corresponding fields.

This process was repeated
1,000 times to estimate the mean proportions of wins for each field,
along with a 95\% confidence interval for these values (Figure~2A).
Additionally, we calculated a Z-score for each field to determine the
number of standard deviations the mean number of wins deviates from
the expected null value of 0.5 for randomly grouped tools (Figure~2B).

@@

%% In brief, software ranks from bioinformatic software benchmarks were
%% gathgered \cite{Gardner:2022}, these cover 68 publications ranking the
%% accuracy 498 distinct software tools in total. Software tools were
%% mapped to departmental field using the address(es) of corresponding
%% authors, these were further mapped to the closest associated ``fields
%% of study'' defined by the National Science Foundation
%% \cite{fields2014}. We analysed fields at three hierarchical levels:
%% \textbf{1.} specific fields (e.g. ``genetics'', ``computer science'',
%% ``bioinformatics''), \textbf{2.} general fields (e.g. ``biological
%% sciences'', ``computer sciences''), \textbf{3.}  expertise area
%% development experts, domain experts or interdisciplinary experts.

%% We then employed a rank-based bootstrapping strategy.  Sets of 200
%% tools are randomly sampled, with replacement, a corresponding
%% benchmark was selected at random for each tool, and the number of
%% times the tool ``won'' against another tool was recorded, along with
%% the total number of pairwise comparisons made. These counts of wins
%% and total comparisons were then assigned to the corresponding
%% specific, general, and expertise areas. This was repeated 1,000 times
%% to estimate the mean proportions of field wins and a 95\% confidence
%% interval (Figure~\ref{fig:fig2}A). Corresponding
%% Z-scores indicate the number of standard deviations
%% from the expected null of 0.5 each field lies (Figure~\ref{fig:fig2}B).

%% $Z-score=\frac{0.5-\mu}{\sigma}$

%% Where $\mu$ is the mean, $\sigma$ is the standard deviation, $x$ is the raw
%% value. P-values are computed from the absolute value of the Z-scores,
%% $P[X > x]$. P-values are corrected for multiple testing by
%% controlling the false discovery rate method
%% \cite{benjamini1995controlling}.

\subsection*{Acknowledgements}

  This research was supported by the MBIE data
science platform ``Beyond Prediction: explanatory and transparent data
science'', Genomics Aotearoa, and Marsden
Grants 19-UOO-040 and 20-UOA-282.

Associate Professor Anthony Gitter (Department of Biostatistics and
Medical Informatics, University of Wisconsin-Madison) for pointing out
the over-representation of Harvard Medical School affiliations in the
Medical Informatics catagory.

% Bibliography
\bibliographystyle{unsrt}
\bibliography{references.bib}
%pdflatex manuscript.tex && bibtex  ./manuscript  && pdflatex manuscript.tex && pdflatex manuscript.tex
